---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# openneuro

<!-- badges: start -->
[![R-CMD-check](https://github.com/opencode-ai/openneuro-r/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/opencode-ai/openneuro-r/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

Search, explore, and download datasets from [OpenNeuro](https://openneuro.org/)
using its GraphQL API. Returns tidy tibbles, supports BIDS-aware file indexing,
and handles selective downloads with local caching.

## Installation

```r
# install.packages("remotes")
remotes::install_github("opencode-ai/openneuro-r")
```

## Usage

### Search for datasets

```{r search, eval = FALSE}
library(openneuro)

search_datasets(modality = "mri", n = 5)
#> # A tibble: 5 × 12
#>   id       name     public version subjects modalities tasks  dataset_name
#>   <chr>    <chr>    <lgl>  <chr>      <int> <list>     <list> <chr>
#> 1 ds008064 Individ… TRUE   1.0.0         28 <list [1]> <list> Individuali…
#> 2 ds008055 Scienti… TRUE   1.0.0        108 <list [1]> <list> Scientific …
#> 3 ds008039 learnin… TRUE   1.0.0         70 <list [1]> <list> learning-ha…
#> 4 ds008034 Action … TRUE   1.0.0         34 <list [1]> <list> Action Repr…
#> 5 ds008018 Personal… TRUE   1.0.0         38 <list [1]> <list> Personality…
```

Filter by task, species, diagnosis, subject count, or author:

```{r search-advanced, eval = FALSE}
search_datasets(
  modality  = "eeg",
  task      = "stroop",
  min_subjects = 30
)
```

### Get dataset metadata

```{r metadata, eval = FALSE}
meta <- get_metadata("ds000001")
meta$title
#> [1] "Balloon Analog Risk-taking Task"
meta$n_subjects
#> [1] 16
meta$tasks
#> [1] "balloon analog risk task"
meta$readme |> substr(1, 200)
#> [1] "This dataset was obtained from the OpenfMRI project
#>      (http://www.openfmri.org).\nAccession #: ds000001\n..."
```

### List files

```{r list-files, eval = FALSE}
# Top-level directory
list_files("ds000001")
#> # A tibble: 23 × 6
#>   id           filename               size directory annexed url
#>   <chr>        <chr>                 <dbl> <lgl>     <lgl>   <chr>
#> 1 bd0bead8ffa… CHANGES                 286 FALSE     FALSE   https://…
#> 2 b6aff27bdd1… README                 1175 FALSE     FALSE   https://…
#> 3 17b93025e53… dataset_description…   615 FALSE     FALSE   https://…
#> 4 436793806e0… participants.tsv       216 FALSE     FALSE   https://…
#> 5 db72baeb366… sub-01                    0 TRUE      FALSE   NA

# Full recursive listing
all_files <- list_files("ds000001", recursive = TRUE)
all_files
#> # A tibble: 136 × 6
```

### Download files

Download entire datasets or just specific subjects:

```{r download, eval = FALSE}
download_dataset("ds000001", subjects = c("01", "02"),
                 dest_dir = "data")
```

Or pick individual files using dplyr-style filtering:

```{r download-filtered, eval = FALSE}
all_files <- list_files("ds000001", recursive = TRUE)

t1w <- all_files[grep("T1w", all_files$filename), ]
download_files(t1w, dest_dir = "data", dataset_id = "ds000001")
```

### BIDS-aware file indexing

Parse BIDS entities from file listings:

```{r bids, eval = FALSE}
idx <- bids_index(all_files)
idx[idx$subject == "01" & idx$suffix == "bold", ]
#> # A tibble: 4 × 11
#>   filename           subject session task   suffix extension
#>   <chr>              <chr>   <chr>   <chr>  <chr>  <chr>
#> 1 sub-01/func/sub-0… 01      NA      ballo… bold   .nii.gz
#> 2 sub-01/func/sub-0… 01      NA      ballo… events .tsv
#> 3 sub-01/func/sub-0… 01      NA      ballo… bold   .nii.gz
#> 4 sub-01/func/sub-0… 01      NA      ballo… events .tsv
```

### Manage the cache

```{r cache, eval = FALSE}
openneuro_cache(info = TRUE)
#> OpenNeuro cache directory: /home/user/.cache/R/openneuro
#> > 0 cached files
#> > Total size: 0 bytes

openneuro_cache(path = "~/my_openneuro_data")
```

## API

| Function | What it does |
|---|---|
| `search_datasets()` | Query by modality, species, task, diagnosis, subject count, author, etc. |
| `get_metadata()` | Dataset title, DOI, subjects, tasks, readme, grant info, validation |
| `list_files()` | Recursive file tree with size, annex status, and download URLs |
| `download_files()` | Download specific files by URL list |
| `download_dataset()` | Download full dataset — optionally filter by subject |
| `bids_index()` | Parse BIDS entities (subject, session, task, run, suffix) from filenames |
| `openneuro_cache()` | Inspect, set, or clear the local download cache |

## Notes

- The `advancedSearch` GraphQL endpoint is used for search. Public datasets that
  produce permission errors are silently skipped with a warning.
- File downloads use the S3 URLs returned by the API. Versioned S3 objects are
  handled automatically.
- The local cache directory defaults to `tools::R_user_dir("openneuro", "cache")`
  and can be overridden with the `OPENNEURO_CACHE_DIR` environment variable.
